---
title: "Project_STAT_Final"
output:
  word_document: default
  pdf_document: default
date: "2024-04-02"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
# Load the dataset
county <- read.csv("county.csv", h = T)

#Essential libraries
library(ggplot2)
library(gridExtra)
library(rpart)
library(Hmisc)
library(usmap)
library(maps)
library(mapproj)
library(tidyverse)
library(dplyr)
library(tidyr)
library(caret)

# View the first few rows of the dataset
head(county)

# Summary statistics for numeric variables
summary(county)

# Structure of the dataset
str(county)

#Now calculate missing values
county <- na.omit(county)

#Scatter plot of poverty VS median_hh_income
ggplot(county, aes(y=poverty, x=median_hh_income))+geom_boxplot()

#In this boxplot without log transformation, the relationship between median_hh_income and poverty appears to be nonlinear and skewed due to the wide range of poverty percentages.

#Use log for better result
ggplot(county, aes(y=log(poverty), x=median_hh_income))+geom_boxplot()

#This plot provides a clearer understanding of how changes in median_hh_income relate to changes in the percentage of the population in poverty, especially for lower income counties where poverty rates are typically higher
```



```{r}
# Load necessary libraries
library(ggplot2)
library(maps)

# Get state data and arrests data
state <- map_data("county") 
poverty <- county  # Assuming "county" contains the poverty data
poverty <- poverty[, c("state", "poverty")]  # Selecting only the "state" and "poverty" columns

# Adjust case for matching 
poverty$region <- tolower(poverty$state)
names(state) <- tolower(names(state))

# Merge and sort (plots in order, sort ensures states filled in)
poverty.geo <- merge(state, poverty, sort = FALSE, by = "region")
poverty.geo <- poverty.geo[order(poverty.geo$order), ]

# Plot
ggplot(poverty.geo, aes(long, lat)) +
  geom_polygon(aes(group = group, fill = poverty)) +
  geom_text(data = poverty.geo %>% group_by(region) %>% summarize(long = mean(long), lat = mean(lat)), 
            aes(label = substr(region, 1, 2)), size = 3, color = "black", fontface = "bold") +  # Add state abbreviations
  coord_map() +
  labs(title = "Poverty Rate by State") +
  scale_fill_viridis_c(name = "Poverty Rate") +
  theme_minimal()


```


```{r}
#Correlation between unemployment rate and median household income
correlation <- cor(county$unemployment_rate, county$median_hh_income)
correlation
#A negative correlation implies that as unemployment rate increases, the median HH income tends to decrease.

ggplot(county, aes(x = log(unemployment_rate), y = log(median_hh_income))) +
  geom_point(color = "blue") +
  geom_smooth(method = "lm", se = FALSE, color = "red") +
  labs(title = "Unemployment Rate vs. Median Household Income",
       x = "Unemployment Rate", y = "Median Household Income") +
  theme_minimal()
#This implies that unemployment rate increases, then median HH income decreases.

```

```{r}

# Homeownership rate comparison between metropolitan and non-metropolitan areas
ggplot(county, aes(x = factor(metro), y = homeownership, fill = metro)) +
  geom_boxplot() +
  labs(title = "Comparison of Homeownership Rate in Metropolitan vs. Non-Metropolitan Areas",
       x = "Metropolitan Area", y = "Homeownership Rate", fill = "Metropolitan Area") +
  theme_minimal()

#The median line is lower for metropolitan areas, it suggests that, on average,non-metropolitan areas have higher homeownership rates.


```
```{r}
# Plotting

# Example: Scatter plot of per capita income vs. median household income
plot(county$per_capita_income, county$median_hh_income, 
     xlab = "Per Capita Income", ylab = "Median Household Income",
     main = "Per Capita Income vs. Median Household Income")
#Each point represents a county, and its position on the plot corresponds to its per capita income and median household income. Here, counties with higher per capita income also tend to have higher median household income.
```


```{r}
# Calculate the average population change from 2010 to 2017
avg_pop_change <- mean(county$pop_change, na.rm = TRUE)
avg_pop_change
#From 2010 to 2017 the average change in population is 58.4%

# Convert 'poverty' variable to numeric
county$poverty <- as.numeric(as.character(county$poverty))

# Filter counties with poverty rate greater than 20%
high_poverty_counties <- county %>% filter(poverty > 12)

#If poverty rate greater than 12% and assigns them to the variable high_poverty_counties


```
```{r}
t1 <- lm( log(poverty) ~ metro, data = county)
summary(t1)

t2 <- lm( log(poverty) ~ median_edu, data = county)
summary(t2)

t3 <- lm( log(poverty) ~ median_hh_income, data = county)
summary(t3)

t4 <- lm( log(poverty) ~ unemployment_rate, data = county)
summary(t4)

t5 <- lm( log(poverty) ~ pop_change, data = county)
summary(t5)

t6 <- lm( log(poverty) ~ homeownership, data = county)
summary(t6)

t7 <- lm( log(poverty) ~ multi_unit, data = county)
summary(t7)
```
```{r}
# Load necessary libraries
library(ggplot2)

# Fit linear regression models
t1 <- lm(log(poverty) ~ metro, data = county)
t2 <- lm(log(poverty) ~ median_edu, data = county)
t3 <- lm(log(poverty) ~ median_hh_income, data = county)
t4 <- lm(log(poverty) ~ unemployment_rate, data = county)
t5 <- lm(log(poverty) ~ pop_change, data = county)
t6 <- lm(log(poverty) ~ homeownership, data = county)
t7 <- lm(log(poverty) ~ multi_unit, data = county)
t8<-lm(log(poverty)~ per_capita_income, data=county)


# Extract coefficients from the models
coefficients <- c(summary(t1)$coef[2, "Estimate"],
                  summary(t2)$coef[2, "Estimate"],
                  summary(t3)$coef[2, "Estimate"],
                  summary(t4)$coef[2, "Estimate"],
                  summary(t5)$coef[2, "Estimate"],
                  summary(t6)$coef[2, "Estimate"],
                  summary(t7)$coef[2, "Estimate"],
                  summary(t8)$coef[2, "Estimate"])

# Names of the variables
variable_names <- c("Metro", "Median Education", "Median Household Income", "Unemployment Rate",
                    "Population Change", "Homeownership", "Multi-unit", "Per Capita Income")

# Create a data frame for visualization
impact_data <- data.frame(variable = variable_names, coefficient = coefficients)

# Calculate percentage of impact for each variable
impact_data$percentage <- impact_data$coefficient / sum(impact_data$coefficient) * 100
# Define custom colors
custom_colors <- c("pink", "skyblue", "grey", "yellow", "orange","red", "blue","purple")

# Plot pie chart with custom colors
ggplot(impact_data, aes(x = "", y = percentage, fill = variable)) +
  geom_bar(stat = "identity") +
  coord_polar("y", start = 0) +
  labs(title = "Impact of Variables on log(Poverty)",
       fill = "Variable") +
  theme_minimal() +
  theme(legend.position = "right") +
  scale_fill_manual(values = custom_colors)

```


```{r}
# Create a new plotting area with 2 rows and 3 columns
par(mfrow=c(2, 3))

# Histogram of log(poverty)
hist(log(county$poverty), probability = TRUE, main = "Log Poverty", xlab = "Log Poverty")

# Histogram of median household income with log transformation
hist(log(county$median_hh_income), probability = TRUE, main = "Median HH Income (log)", xlab = "Log Median HH Income")

# Histogram of unemployment rate with log transformation
hist(log(county$unemployment_rate), probability = TRUE, main = "Unemployment Rate (log)", xlab = "Log Unemployment Rate")

# Histogram of population change with log transformation
hist(log(county$pop_change), probability = TRUE, main = "Population Change (log)", xlab = "Log Population Change")

# Histogram of homeownership with log transformation
hist(log(county$homeownership), probability = TRUE, main = "Homeownership (log)", xlab = "Log Homeownership")

# Histogram of per capita income with log transformation
hist(log(county$per_capita_income), probability = TRUE, main = "Per Capita Income (log)", xlab = "Log Per Capita Income")

# Reset the plotting area
par(mfrow=c(1, 1))


```


```{r}

# Example: Bar plot of unemployment rate by state
# Load necessary libraries
library(ggplot2)
library(dplyr)

# Calculate average unemployment rate by state
unemployment_by_state <- county %>% 
  group_by(state) %>% 
  summarize(avg_unemployment_rate = mean(unemployment_rate, na.rm = TRUE))

# Plot the bar plot
ggplot(unemployment_by_state, aes(x = state, y = avg_unemployment_rate)) +
  geom_bar(stat = "identity", fill = "skyblue") +
  labs(title = "Average Unemployment Rate by State",
       x = "State",
       y = "Average Unemployment Rate") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))  # Rotate x-axis labels for better readability
```
```{r}

# Load necessary libraries
library(ggplot2)
library(maps)

# Check if the 'unemployment' variable is present in the 'unemployment.geo' data frame
# "unemployment" %in% names(unemployment.geo)

# Get state data and unemployment data
state <- map_data("county") 
unemployment <- county 
unemployment <- unemployment[, c("state", "unemployment_rate")]  

# Adjust case for matching 
unemployment$region <- tolower(unemployment$state)
names(state) <- tolower(names(state))

# Merge and sort (plots in order, sort ensures states filled in)
unemployment.geo <- merge(state, unemployment, sort = FALSE, by = "region")
unemployment.geo <- unemployment.geo[order(unemployment.geo$order), ]

# Plot
ggplot(unemployment.geo, aes(long, lat)) +
  geom_polygon(aes(group = group, fill = unemployment_rate)) +
  geom_text(data = unemployment.geo %>% group_by(region) %>% summarize(long = mean(long), lat = mean(lat)), 
            aes(label = substr(region, 1, 2)), size = 3, color = "black", fontface = "bold") +  # Add state abbreviations
  coord_map() +
  labs(title = "Unemployment Rate by State") +
  scale_fill_viridis_c(name = "Unemployment Rate") +
  theme_minimal()

#As the shade gets lighter the poverty rate of that state increases. And as the shade gets darker the poverty rate in that state is low.
```


#Research Questions
#1. Variability of poverty rate across different states
#2. What is the Correlation between unemployment rate and state?
#3. What is the relationship between level of per capita income and median_hh_income?
#4. Compare the homeownership rate between metropolitan and non-metropolitan areas
#5. What is the average population change from 2010 to 2017?

```{r}
# Interaction between variables
model_lm_1 <- lm( log(poverty)~median_edu , data = county)
model_lm_1
#here 71.428%---->((3.6-2.1)/2.1) of population is having more poverty rate as compared to bachelors. Bachelors are having less poverty rate compared to below_hs, hs_diploma and some_college
summary(model_lm_1)
anova(model_lm_1)
#MSE=0.130, R2_adj=0.2398

model_lm_2 <- lm( log(poverty)~median_edu+median_hh_income , data = county)
model_lm_2
summary(model_lm_2)
anova(model_lm_2)
#MSE=0.058, R2_adj=0.6604

model_lm_3 <- lm( log(poverty)~median_edu + median_hh_income + unemployment_rate , data = county)
model_lm_3
summary(model_lm_3)
anova(model_lm_3)
#MSE=0.051, R2_adj=0.6999

model_lm_4 <- lm( log(poverty)~median_edu + median_hh_income + unemployment_rate + metro , data = county)
model_lm_4
summary(model_lm_4)
anova(model_lm_4)
#MSE=0.049, R2_adj=0.7171

model_lm_5<-lm(log(poverty)~median_edu + median_hh_income + unemployment_rate + metro + per_capita_income, data = county)
model_lm_5
summary(model_lm_5)
anova(model_lm_5)

ggplot(county, aes(x=median_edu, y=poverty, color=metro))+geom_boxplot()


R2_adjusted<-c(0.2398,0.6604,0.6999, 0.7171, 0.7281 )
R2_adjusted
MSE<-c(0.130,0.058,0.051,0.049,0.047)
MSE
#From above all 4 models, MSE has been decreasing and R2_adj is increasing for addition of variables.

#As we move from Model_lm_1 to Model_lm_4, we see improvements in both MSE and Adjusted R-squared values, indicating better predictive performance and explanatory power
```
# Logistic regression models are evaluated based on metrics such as accuracy, precision, recall, and ROC curves, which are suitable for binary classification tasks.Multiple linear regression models are evaluated based on metrics such as MSE and R-squared, which are more appropriate for numerical outcomes.

# Here we dont have continuous outcomes thats why we converted the data into binary and we used logistics regression for more accuracy.

```{r}
#Add poverty_high variable to the dataset
# Load the dataset
data(county)

# Preprocess the data
# Remove NA values
county <- na.omit(county)

# Convert poverty into a binary variable
county$poverty_high <- ifelse(county$poverty > median(county$poverty), 1, 0)

# Encode states as factors
county$state <- as.factor(county$state)

# Perform logistic regression
model <- glm(poverty_high ~ median_edu + median_hh_income + unemployment_rate + metro+ per_capita_income, data = county, family = "binomial")

# Interpret the results
summary(model)

View(county)
```
#We ran a logistic regression model to predict whether a county has a high poverty rate based on several predictors: median_edu,median_hh_income, unemployment_rate, and whether the county is metropolitan (metro)

# The model shows that higher education and higher household incomes gives lower odds of poverty rate. Also presence of metropolitan area appears to be associated with lower odds of high poverty.

#This suggests that efforts to improve education, increase household incomes, and address unemployment could contribute to reducing high poverty rates in counties.



```{r}
#Prediction Accuracy Using Confusion matrix
# Load required libraries
library(dplyr)
library(tidyr)
library(caret)

# Load the dataset
data(county)
n<-dim(county)
n

# Preprocess the data
# Remove NA values
county <- na.omit(county)

# Convert poverty into a binary variable
county$poverty_high <- ifelse(county$poverty > median(county$poverty), 1, 0)

# Encode states as factors
county$state <- as.factor(county$state)


N<-dim(county)
N
set.seed(123)
train_index<-sample(1:2560, 0.8*2560)
train_data<-county[train_index,]
test_data<-county[-train_index,]
# Perform logistic regression
model <- glm(poverty_high ~ median_edu + median_hh_income + unemployment_rate + metro+per_capita_income, data = train_data, family = "binomial")

# Make predictions on testing data
predictions <- ifelse(predict(model, newdata = test_data, type = "response") > 0.6, 1, 0)

# Create confusion matrix
confusion_matrix <- table(Actual = test_data$poverty_high, Predicted = predictions)
confusion_matrix

# Calculate accuracy
accuracy <- sum(diag(confusion_matrix)) / sum(confusion_matrix)
accuracy


# Using class technique
prob1<-stats::predict(model, newdata=test_data)#  this will result in log odds ratio
prob1<-stats::predict(model, newdata=test_data, type="response") 
# `type=response` provides predicted probability 


pred1<- rep(0,dim(test_data)[1]) ## create a zero vector, there are 2048 rows  in the training set

pred1[prob1>0.6]=1 
tab1<-table(pred1, test_data$poverty_high) # confusion matrix
addmargins(tab1)
 sum(diag(tab1))/sum(tab1)

```

#The logistic regression model trained on the dataset achieved an overall accuracy of approximately 82.42% in predicting whether a county has a high poverty rate.

```{r}
plot_confusion_matrix <- function(model, test_data, lower_prob = 0.1, upper_prob = 1, threshold_step = 0.1) {
  # Initialize empty lists to store results
  accuracy_values <- c()
  thresholds <- seq(lower_prob, upper_prob, by = threshold_step)
  
  # Iterate over different probability thresholds
  for (threshold in thresholds) {
    # Predict outcomes using the given threshold
    predictions <- ifelse(predict(model, newdata = test_data, type = "response") > threshold, 1, 0)
    
    # Create confusion matrix
    confusion_matrix <- table(Actual = test_data$poverty_high, Predicted = predictions)
    
    # Calculate accuracy and store in list
    accuracy <- sum(diag(confusion_matrix)) / sum(confusion_matrix)
    accuracy_values <- c(accuracy_values, accuracy)
  }
  
  # Plot accuracy versus probability thresholds
  plot(thresholds, accuracy_values, type = "l", 
       xlab = "Probability Threshold", ylab = "Accuracy", 
       main = "Accuracy vs. Probability Threshold")
}

plot_confusion_matrix(model, test_data, lower_prob = 0.1, upper_prob = 1, threshold_step = 0.1)


```

```{r}
# Load necessary libraries
library(tidyverse)

# Load the dataset
data(county)

# Fit gamma regression model
gamma_model <- glm(poverty ~  median_edu + median_hh_income + unemployment_rate + metro+per_capita_income , data = county, family = "Gamma")

# Summarize the model
summary(gamma_model)
anova(gamma_model)

```





